To be FAIR: Theory Specification Needs an Update

### 2025-03-02 manuscript.Rmd
Move paragraph about science communication to future directions, #65

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,4"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "5"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Noah N.N. van Dongen"
    affiliation: "6"
    role:
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,7"
    role:
      - Writing – review & editing
      - Investigation
affiliation:
  - id            : "1"
    institution   : "Tilburg University, dept. Methodology & Statistics"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
  - id            : "5"
    institution   : "Ludwig-Maximilians-Universität München, Germany"
  - id            : "6"
    institution   : "University of Amsterdam, the Netherlands"
  - id            : "7"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`.

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

@@

The so-called "replication crisis" has prompted extensive reforms in psychology and other scientific fields [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypotheses tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.

@@

Scholars have been raising concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One main concern is that theories lack or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
<!-- I still believe formalization != precision. Theories can be 100% formal and still 0% precise. -->
<!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning). However, what we mean, I think, is the explication of a narrative in (something so detailed and explicit that it can be captured with little error in) analytical or computational form (but the meaning remains. E, M, and C stand form something is E = MC<sup>2</sup>).   -->
When theories are difficult to understand on their own,
without either substantial interpretation or additional background knowledge,
it becomes difficult do derive precise predictions,
and therefore hard to falsify the theory. 
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.

@@

<!-- We don't explicitly define theory, but discuss the most important ideas of what theory might be, and advocate for a pluralism of information that can be tracked as FAIR theory (e.g. also such things as scientific practice (experimental design, materials)). -->
<!-- The nature of the tracked information can change over time, as a scientific theory matures. -->
<!-- The discussion about the distinction between FAIR theories and models is removed, as I believe it might be confusing for first-time readers and not really necessary for the bigger picture of the paper (but of course could be re-added). -->
<!-- NvD: I wonder if we need what is stated about the syntactic, semantic, and pragmatic perspectives. We can leave it at the characterization that theories are (typically) concerned with the explanation and understanding of what can be observed (directly or indirectly) tested (in the laboratory or natural experiments) and manipulated. But, this can come in many forms and we do not exclude the possibility that there are acceptions to this characterization that we could still consider a scientific theory. We can then state that we use Meehl's view for convenience in this paper, but that it is compatible other perspectives (or something.  -->
Definitions of theory abound and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
the method is not contained to any one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's hierarchy of ever-more specific "statements" a theory might contain [-@meehlAppraisingAmendingTheories1990]:
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].

@@

Findability is substantially amplified if intended users of a resource know where to search for it.
This is a known problem in relation to research data and software [@katzSpecialIssueSoftware2024].
Regrettably, most academic search engines only index traditional print publications, not other information artifacts.
Since the status quo is to publish theories in papers,
the FAIR requirements are met if scholars continue to do so,
and additionally publish theories as separate information artifacts.
The `"fairtheory"` keyword can also be used to signal the presence of theory within a paper.
<!-- An ad-hoc solution is to publish a paper as documentation for the information artifact [@mcgillivrayDeepImpactStudy2022]. -->
<!-- There is even a journal - the [Journal of Open Source Software](https://joss.theoj.org/) - that publishes software documentation as a paper so that it will be indexed by search engines. -->
<!-- The same solution could be applied to theories. -->
In the longer term, it may not be necessary to write a paper for each theory.
If Zenodo becomes more recognized as a centralized repository for information artifacts, researchers may begin to search there more regularly.
Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index information artifacts stored in Zenodo.

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [@grayHowMapTheory2017].
Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
<!-- Is there something to reference? Website/preprint? Otherwise is seems a bit weird to just mention this without any further information. -->
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
for example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

The FAIR theory workflow documented above offers several concrete ways to make changes to a FAIR theory object.
If you start with an existing GitHub repository, and wish to make some changes to it,
this is commonly done by [creating a "branch"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-and-deleting-branches-within-your-repository).
A branch allows you to make non-destructive changes and can continue to exist in parallel to the main repository.
Thus, it is possible to have one main theory with several branches that each contain competing theories derived from it.
Figure \@ref(fig:figbranch) shows how our example `empirical_cycle` repository contains branches with Wagenmakers' and Van Lissa's implementations.
A branch can also be merged with the main branch, thus incorporating the changes it contains into the version of record.
If you wish to develop a new version of someone else's FAIR theory,
it is possible to ["fork their repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo).
<!-- I think it would be nice if links are colored (blue), so people can see in the PDF that there is something to click. -->
This creates a copy of their repository onto your GitHub account.
Both branches and forks can be compared and merged via "pull requests", which are essentially a request to incorporate the changes you have made.
Figure \@ref(fig:figcompare) shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.

@@

Finally, FAIR theory holds substantial potential for science communication. While the general public is ill-equipped to read and derive actionable insight from large quantities of empirical papers about a particular phenomenon,
theories encapsulate the bigger picture of contemporary scientific understanding.
For example, while few people read empirical studies on attachment,
attachment theory plays a prominent role in popular scientific books about parenting and romantic relationships.
Theory bridges the gap between academic research and practitioners by summarizing actionable insights, relieving practitioners from the need to sift through extensive empirical literature.
By providing a mechanism for iterative improvement based on emerging evidence, FAIR theory also supports effective evidence-based decision making.

@@

@@

FAIR theory is a major step forward towards more transparent, collaborative, and efficient theory construction.
It provides much-needed open science methods for the inductive phase of the empirical cycle,
closing a critical gap in the scientific process.
FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
This paves the way for more theory-driven scholarship,
accelerating cumulative knowledge acquisition in psychology, other social sciences, and beyond.

>>>>

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,4"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "5"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Noah N.N. van Dongen"
    affiliation: "6"
    role:
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, dept. Methodology & Statistics"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
  - id            : "5"
    institution   : "Ludwig-Maximilians-Universität München, Germany"
  - id            : "6"
    institution   : "University of Amsterdam, the Netherlands"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`.

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences, and beyond.

@@

The so-called "replication crisis" has prompted extensive reforms in psychology and other scientific fields [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
Under the original definition, few would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience
Yet, under the expanded framework, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.

@@

Scholars have been raising concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One main concern is that theories lack precision, or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
<!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning). However, what we mean, I think, is the explication of a narrative in (something so detailed and explicit that it can be captured with little error in) analytical or computational form (but the meaning remains. E, M, and C stand form something is E = MC<sup>2</sup>).   -->
When theories do not make precise predictions,
they are hard to falsify and difficult to understand on their own,
without either substantial interpretation or additional background knowledge.
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.

@@

<!-- We don't explicitly define theory, but discuss the most important ideas of what theory might be, and advocate for a pluralism of information that can be tracked as FAIR theory (e.g. also such things as scientific practice (experimental design, materials)). -->
<!-- The nature of the tracked information can change over time, as a scientific theory matures. -->
<!-- The discussion about the distinction between FAIR theories and models is removed, as I believe it might be confusing for first-time readers and not really necessary for the bigger picture of the paper (but of course could be re-added). -->
<!-- NvD: I wonder if we need what is stated about the syntactic, semantic, and pragmatic perspectives. We can leave it at the characterization that theories are (typically) concerned with the explanation and understanding of what can be observed (directly or indirectly) tested (in the laboratory or natural experiments) and manipulated. But, this can come in many forms and we do not exclude the possibility that there are acceptions to this characterization that we could still consider a scientific theory. We can then state that we use Meehl's view for convenience in this paper, but that it is compatible other perspectives (or something.  -->
Definitions of theory abound and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
the method is not contained to any one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories]
The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's hierarchy of ever-more specific "statements" a theory might contain [-@meehlAppraisingAmendingTheories1990]:
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].

@@

Findability is substantially amplified if intended users of a resource know where to search for it.
This is a known problem in relation to research data and software [@katzSpecialIssueSoftware2024].
Regrettably, most academic search engines only index traditional print publications, not other information artifacts.
Since the status quo is to publish theories in papers,
the FAIR requirements are met if scholars continue to do so,
and additionally publish theories as separate information artifacts.
The `"fairtheory"` keyword can also be used to signal the presence of theory within a paper.
<!-- An ad-hoc solution is to publish a paper as documentation for the information artifact [@mcgillivrayDeepImpactStudy2022]. -->
<!-- There is even a journal - the [Journal of Open Source Software](https://joss.theoj.org/) - that publishes software documentation as a paper so that it will be indexed by search engines. -->
<!-- The same solution could be applied to theories. -->
In the longer term, it may not be necessary to write a paper for each theory.
If Zenodo becomes more recognized as centralized repository for information artifacts, researchers may begin to search there more regularly.
Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index information artifacts stored in Zenodo.

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [@grayHowMapTheory2017].
Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
for example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

The FAIR theory workflow documented above offers several concrete ways to make changes to a FAIR theory object.
If you start with an existing GitHub repository, and wish to make some changes to it,
this is commonly done by [creating a "branch"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-and-deleting-branches-within-your-repository).
A branch allows you to make non-destructive changes and can continue to exist in parallel to the main repository.
Thus, it is possible to have one main theory with several branches that each contain competing theories derived from it.
Figure \@ref(fig:figbranch) shows how our example `empirical_cycle` repository contains branches with Wagenmakers' and Van Lissa's implementations.
A branch can also be merged with the main branch, thus incorporating the changes it contains into the version of record.
If you wish to develop a new version of someone else's FAIR theory,
it is possible to ["fork their repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo).
This creates a copy of their repository onto your GitHub account.
Both branches and forks can be compared and merged via "pull requests", which are essentially a request to incorporate the changes you have made.
Figure \@ref(fig:figcompare) shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.

@@

@@

Another future direction for FAIR theory is as an instrument of science communication. Practitioners and the general public are rarely able to read and derive actionable insight from large quantities of empirical papers about a particular phenomenon.
Theories are more accessible, because they encapsulate the bigger picture of contemporary scientific understanding.
For example, while few people read empirical studies on attachment,
attachment theory plays a prominent role in popular scientific books about parenting and romantic relationships.
Theory bridges the gap between academic research and practitioners by summarizing actionable insights, relieving practitioners from the need to sift through extensive empirical literature.
By providing a mechanism for iterative improvement based on emerging evidence, FAIR theory also supports effective evidence-based decision making.

@@

FAIR theory is a major step forward towards more transparent, collaborative, and efficient theory construction.
It provides much-needed open science methods for the inductive phase of the empirical cycle,
closing a critical gap in the scientific process.
FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
This paves the way for more theory-driven scholarship,
and accelerates cumulative knowledge acquisition in the psychology, other social sciences, and beyond.
